---
title: "HW4_Paces_Will"
author: "Will Paces"
date: "March 9, 2022"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## STAA 566 Will Paces Homework 4 - Tables

I wanted to visualize the change in state populations between 2010 and 2019, and provide plots detailing the change in median income over this time period in an attempt to provide an economic explanation for the migration of people. This data was downloaded from the US census bureau from both the Decennial and American Community Survey data repositiories. I decided to visualize this data in a table with spark charts detailing the trend in both population and median income growth metrics in order to allow users to make visual comparisons between these trends, as well as see overall changes during this timeframe. 

```{r, warning=FALSE,message=FALSE}
# Load libraries
library(tidycensus)
library(tidyverse)
library(tidyr)
library(dplyr)
library(rlang)
library(purrr)
library(sparkline)
sparkline(0)
library(reshape2)
library(kableExtra)
library(knitr)
```

```{r}
# Get 2010 population data, save state name, fips and population columns
statepop.2010 <- get_decennial(geography = 'state', variables = 'P001001', year = 2010)
statepop.2010 <- statepop.2010 %>% select(NAME, fips = GEOID, pop2010 = value)
statepop.2010$fips <- as.integer(statepop.2010$fips)

# Get 2019 (2020 is unavailable?) population data, save fips and pop
statepop.2019 <- get_acs(geography = 'state', variables = 'B01001_001', year = 2019)
statepop.2019 <- statepop.2019 %>% select(fips = GEOID, pop2019 = estimate)
statepop.2019$fips <- as.integer(statepop.2019$fips)

# Merge datasets by fips, calculate population change
statepop.diff <- merge(statepop.2010, statepop.2019, by = 'fips')
statepop.diff$popChange <- statepop.diff$pop2019 - statepop.diff$pop2010
statepop.diff <- statepop.diff %>% select(NAME, popChange)
names(statepop.diff) <- c('state', 'popChange')

# Download median income data per state between 2010 and 2019 ----
years <- c(2010:2019)
state.income <- map_dfr(years, 
                        ~get_acs(geography = 'state', variables = 'B19013_001', 
                                 year = .x, geometry = F),
                        .id = 'year.id') %>%
  select(year.id, NAME, median.income = estimate) %>% # remove unnecessary variables
  mutate(year = as.integer(year.id) + 2009) %>%
  select(-year.id)

state.income <- state.income[order(state.income$NAME), ]

# Compare beginning and final median state income during time period
state.income.delta <- data.frame(matrix(NA, nrow = 0, ncol = 2))
names(state.income.delta) <- c('state', 'income.change')

for(i in unique(state.income$NAME)){
  sub.data.start <- state.income[state.income$NAME == i & state.income$year == 2010, ]
  sub.data.end <- state.income[state.income$NAME == i & state.income$year == 2019, ]
  state.add <- data.frame("state" = i, "income.change" = sub.data.end$median.income - sub.data.start$median.income)
  state.income.delta <- rbind(state.income.delta, state.add)
}

# Load total population estimate for 2010 to 2019 plot ----
# Data was downloaded as a spreadsheet from https://www.census.gov/data/tables/time-series/demo/popest/2010s-state-total.html 
state.pop.plot.data <- read.csv('C:\\Users\\Will\\Desktop\\STAA 566\\AssignmentGitRepos\\HW4_Paces_Will\\USgovCensusStatePopulationTS.csv')
names(state.pop.plot.data) <- c('state', '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017', '2018', '2019')

# convert to long data
state.pop.plot.data <- melt(state.pop.plot.data, id.vars =  'state', variable.name = 'year', value.name = 'population')

# Remove "."s from state name variable
state.pop.plot.data$state <- gsub("\\.", "", state.pop.plot.data$state)

# order by state names and declare names as factor
state.pop.plot.data <- state.pop.plot.data[order(state.pop.plot.data$state), ]


# Ensure that statepop.diff and state.income dataframes have states organized in the same (alphabetical) order
statepop.diff <- statepop.diff[order(statepop.diff$state), ]

# Merge datasets
state.data <- merge(statepop.diff, state.income.delta, by = 'state')

# Establish variable for income change charts
state.data <- state.data %>%
  mutate(plt.income = NA, plt.pop = NA)

# Add chart data to table
for(st in state.data$state){
  
  # add charts for population growth
  state.pop <- state.pop.plot.data %>%
    filter(state == st) %>%
    arrange(year) %>%
    pull(population)
  
  state.data[state.data$state == st, "plt.pop"] <- spk_chr(state.pop, type = 'line',
                                                           chartRangeMin = min(state.pop),
                                                           chartRangeMax = max(state.pop))
  
  # add charts for income growth
  state.inc <- state.income %>%
    filter(NAME == st) %>%
    arrange(year) %>%
    pull(median.income)
  
  state.data[state.data$state == st, "plt.income"] <- spk_chr(state.inc, type = 'line',
                                                              chartRangeMin = min(state.inc),
                                                              chartRangeMax = max(state.inc))
}

# Organize data for table creation
state.data.table <- state.data %>%
  select(state, popChange, plt.pop, income.change, plt.income) %>%
  kbl(escape = F,
      col.names = c("State", "Population Change", "Population Trend",
                    "Median Income Change", 'Median Income Trend'),
      align = c('l','c', 'c', 'c', 'c')) %>%
  kable_styling(bootstrap_options = 'hover', fixed_thead = T) %>%
  add_header_above(c(" ", "2010 to 2019 Time Frame" = 4))
  
state.data.table


```
